\documentclass{article}
\setlength\parindent{0pt}

\begin{document}

\section {Value chain \& information systems}

\subsection{Primary activities}

Inbound logistics (getting resources and checking them to sell)

\begin{itemize}
\item  Example: Delhaize automating a factory, able to deliver fresher products.

\begin{itemize}
\item  Rivalry among existing competition =$>$ Delhaize has fresh products

\item  Threat of New entrants =$>$ Delhaize invested a lot, this discourages new companies to try to setup something equal.

\item  Bargaining power of buyers =$>$ they have the freshest products, so they can raise their price.
\end{itemize}
\end{itemize}

Operations (how to make the product, with manufacturing and maintenance)

\begin{itemize}
\item  Example: Ryanair will only serve their customers through their website
\end{itemize}

Outbound logistics (delivering the product or service)

\begin{itemize}
\item  Example: Amazon wants to use drones to deliver their products, they even claim that they know what you will buy before you even click the order button.
\end{itemize}

Marketing and Sales (promoting and selling products)

\begin{itemize}
\item  Example: Colruyt delivers \underbar{personalized} folders with promotions!
\end{itemize}

Customer service

\begin{itemize}
\item  Example: Booking.com where they ask the costumers for feedback, so they can get a lot of information about the hotels and costumers. 
\end{itemize}

\subsection{Support activities}

HR analytics 

\begin{itemize}
\item  Gaining knowledge about which employee may leave the company

\item  With this information you can attract new employees and create a good environment so your employees don't leave
\end{itemize}

\subsection{Key message}

\begin{itemize}
\item  Alignment between your business model and your IT strategy is needed for your organization to be successful
\end{itemize}

\section{Linking business \& IS strategy}
 
\subsection{Strategic alignment model}

The ICT strategy is created with respect to the external market and what other competitors do.

\hfill \newline
Two models: 

\begin{itemize}
\item  Strategic fit (Business)

\begin{itemize}
\item  How your strategy matches with your business model.
\end{itemize}

\item  Strategic fit (IT)

\begin{itemize}
\item  How your IT strategy matches with your operations
\end{itemize}

\item  Functional fit

\begin{itemize}
\item  Whether your business model and IT strategy are aligned.
\end{itemize}
\end{itemize}

\subsection{Strategic transformation}

\begin{itemize}
\item  First create business model and then setup your ICT operations to support this model
\end{itemize}

\subsection{Technology transformation}

\begin{itemize}
\item  Create an ICT infrastructure to accomplish your business model
\end{itemize}

\subsection{Competitive potential}

\begin{itemize}
\item  What technologies do I already have to create a good business model around it.
\end{itemize}

\subsection{Service level}

\begin{itemize}
\item  Create an ICT strategy, create and develop the system and notice that you can sell this technology to other companies. (You create a business system around it)
\end{itemize}

\section{IT governance}

\subsection{Governance}

Governance is a control system to decide about IT and business strategy and making sure that the strategy is executed.

\hfill \newline
Corporate governance and fraud 

\begin{itemize}
\item  Led to new regulations (SOX)
\end{itemize}

\subsection{Internal control (structure)}

A dynamic process that is effected by the board of directors, designed to provide reasonable assurance regarding the achievement of objectives relating to everything about running a business.

\hfill \newline
Build frameworks that allow organizations to demonstrate that this control system is in place.

\begin{itemize}
\item  COSO internal control (framework)

\begin{itemize}
\item  Demonstrates that the SOX compliance is in place.

\item  No IT
\end{itemize}
\end{itemize}

\subsection{IT governance}

IT governance is the organizational capacity exercised by the board of directors, executive management and IT management to control the formulation and implementation of IT strategy and in this way ensure the fusion of business and IT.

\begin{itemize}
\item  Goal:

\begin{itemize}
\item  Better organizational performance

\item  Better alignment between your business and IT
\end{itemize}
\end{itemize}

COBIT 

\begin{itemize}
\item  show compliance with the SOX act

\item  professional framework
\end{itemize}

COBIT principles

\begin{itemize}
\item  read text Toledo
\end{itemize}

\end{document}

